HMMotion: Using tracking data to predict coverage strategies in American Football

### 2024-12-31 submission.Rmd
.

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns). However, these models seem to be trained on plays without pre-snap motion, or at least to the situations before motions (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), which is a crucial element of modern offensive strategies.

@@

## Data
We analyze tracking data from nine weeks of the NFL 2022 season, provided by the NFL Big Data Bowl 2024. Beside the tracking data, we also use information on plays and players. We further considered the corresponding data from PFF that assigned the categories \textit{man}, \textit{zone} and \textit{others} representing the different schemes to each play. As it is not properly described what \textit{others} means, we omit every play that is associated with this value. Moreover, we omit plays with more than five offensive linemen and with two quarterbacks `r colorize("and those plays that did not contain any pre-snap motion","orange")` (`r colorize("Ich würde deutlicher drauf eingehen, dass wir nur an motion plays interessiert sind! Vl ein Punkt machen und schreiben: Since we are specifically interested in analyzing pre-snap player movements, we omit plays that did not contain any pre-snap motion.","red")`). Then, we end up with 3985 offensive plays in total, from which the defense played 2973 in zone and 1012 in man coverage.

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as covariates and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

>>>>

## Introduction
Deciding between man and zone coverage is one of the most critical strategic choices a defensive coordinator must make before each offensive play in American football. While experienced offensive coordinators and quarterbacks often rely on visual cues to identify these defensive schemes, the increasing availability of player tracking data offers a new avenue to uncover and analyze these tactics. A notable example is Amazon's NFL Next Gen Stats model, which delivers coverage predictions during live broadcasts (see a snapshot of the 2024 Week 12 matchup between the Pittsburgh Steelers and Cleveland Browns). However, these models seem to be trained on plays without pre-snap motion, or at least to the situations before motions (see [Amazon](https://aws.amazon.com/de/blogs/machine-learning/identifying-defense-coverage-schemes-in-nfls-next-gen-stats/)), while motion is a crucial element of modern offensive strategies.

@@

## Data
We analyze tracking data from nine weeks of the NFL 2022 season, provided by the NFL Big Data Bowl 2024. Beside the tracking data, we also use information on plays and players. We further considered the corresponding data from PFF that assigned the categories \textit{man}, \textit{zone} and \textit{others} representing the different schemes to each play. As it is not properly described what \textit{others} means, we omit every play that is associated with this value. Moreover, we omit plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we omit plays that did not contain any pre-snap motion. Then, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage.

##### Feature engineering
To accurately forecast the defensive scheme (man- or zone defense) for every play, we need to create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: We first consider all 11 players on each side of the field and compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, while, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five attackers corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. Finally, we use their standardized $x$ and $y$ coordinates as covariates and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback (`r colorize("values taken at event lineset. Rouven: Die Values haben wir doch rausgenommen aus dem finalen Modell oder?","orange")`). Finally, we extract relevant information from play-by-play data, such as quarter, down, yards to go, home and away score and the game clock.

@@

We locally decoded this sequence.
